---
title: "Problem Set 6"
author: "Sara Peters"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(kableExtra)
library(tinytex)
library(knitr)

#Set working directory
setwd("C:/Users/saraa/OneDrive - UW-Madison/SOC 756- Demography Techniques II/Problem Sets/DemTech2/Problem Set 6")

#load libraries
# install.packages("dplyr")
# install.packages("tidyverse")
# install.packages("ggplot2")
#install.packages("readxl")


library(dplyr)
library(tidyverse)
library(ggplot2)
#library (readxl)

# Set the 'scipen' option to a large value to prevent scientific notation
options(scipen = 300)

# Set the 'digits' option to control the number of decimal places
options(digits = 6)  # Change the number to the desired decimal places

```


###Soc 756

#Multiregional Projection

#This exercise requires that you adapt the model of Mare's 1997 "Differential Fertility…" article to the specific constraints and possibilities of the data supplied below. 

#You have the data given below in Tables 1-4 for black and white women for a given year. (Percentage totals in these tables may not sum to 100 because of rounding.)


#For this exercise, assume that (1) the population is closed to immigration and emigration, (2) the population has only one sex (females) and two race categories, (3) all fertility is confined between exact ages 15 and 45 and age-specific fertility rates are constant within this interval, and (4) fertility, mortality, and mobility rates are fixed over time. 

#Several of the questions below ask you to describe the equilibrium education distributions. This assumes that there is an equilibrium distribution of persons in the P(t+1) vector across education groups associated with the Leslie matrix and the starting P(0) vector. That is, repeated application of the Leslie matrix to the P(0) vector will eventually produce a P(n) vector in which the distribution of persons across education groups is not changing with each subsequent multiplication.



#This problem set will be submitted in two parts:


##Generate an M matrix with data for Black women and an M matrix with data for White women. You may have to make some decisions in constructing these. Make note of these for the answer to question number 1, to be submitted the following week. Upload your R script that constructs the two matrices to your GitHub page. The only “submission” this week to share the link to your uploaded file to me and two people in the class.

#Assumptions: 

1. Everyone exits the population by age 50
2. There is no intragenerational educational mobility, ultimate educational attainment is determined at birth
3. Fertility below age 15 and above age 45 are zero
4. Fertility, mortality, and mobility are constant
5. Initial population is distributed into education groups based on % Distribution of educational attainment for black and white women (Table 1). 


***1. What decisions did you have to make to work with the data that you received for this assignment? Do these decisions require assumptions?***

I had to make decisions about the mortality and fertility patterns of the population and how to distribute the initial population to run the simmulation. All of these decisions require assumptions. For example, I assume all births happen between ages 15 and 45, and that all members of the population exit by age 50. 


***2. What equilibrium distributions of educational attainment for Black and White people are implied by the data in Tables 1-4?***


***3. To what degree are differences in the (equilibrium) education distributions for Black and White people attributable to differences in fertility patterns between Black and White women?***
***4. To what degree are differences in the (equilibrium) education distributions for Black and White people attributable to differences in maternal mortality patterns between Black and White women?***
***5. To what degree are differences in the (equilibrium) education distributions for Black and White people attributable to differences in child (under 15) mortality patterns by race?***
***6. To what degree are differences in the education distributions for Black and White people after three generations attributable to differences in intergenerational mobility patterns between Black and White people?***
***7. What would be the equilibrium education distributions for Black and White people if there were NO intergenerational educational mobility?***
***8. What would the equilibrium distributions be if mother's and daughter's educational attainments were statistically independent?***
***9. Extra credit: Choose one of the questions above and plot the population educational distribution over time to determine (a) whether equilibria exist and (b) how long it takes to achieve.***







##Create Dataframes to supply matrices
```{r}

Lx_fem<- read.csv("nLx_values.csv")

#create separate black and white survival dataframes
Lx_fem$group <- substr(Lx_fem$cat_race_ed_age,1,3)

Lx_fem$race <- substr(Lx_fem$cat_race_ed_age, 2,2)

Lx_fem$age <- gsub(".*_", "", Lx_fem$cat_race_ed_age)

black_fem <- Lx_fem[which(Lx_fem$race=="b"),]

white_fem <- Lx_fem[which(Lx_fem$race=="w"),]

#create master mobility dataframes from Table 3 in mare
black_mobility <-data.frame(
  edu = c(8, 11, 12, 15, 16),
  black_edu = c(37.6, 29.0, 23.3, 6.3, 3.8), 
  black_em_8 = c(28.9, 7.9, 2.5, 3.3, 0.0), 
  black_em_11 = c(26.8, 35.0, 19.0, 3.8, 3.2), 
  black_em_12 = c(24.3, 27.8, 38.6, 24.3, 16.3), 
  black_em_15 = c(12.6, 16.4, 21.2, 49.6, 37.1), 
  black_em_16 = c(7.3, 12.9, 18.8, 18.9, 43.4) 
 
)

white_mobility <-data.frame(
  edu = c(8, 11, 12, 15, 16),
  white_edu = c(17.2, 21.9, 42.1, 11.9, 6.9),
  white_em_8 = c(13.2, 3.4, 1.4, 1.0, 0.1), 
  white_em_11 = c(17.9, 15.0, 6.1, 2.7, 3.3), 
  white_em_12 = c(48.5, 42.7, 45.7, 24.3, 14.3), 
  white_em_15 = c(13.0, 20.8, 24.0, 33.8, 25.9), 
  white_em_16 = c(7.5, 18.0, 21.7, 38.1, 56.4) 
)

#subset to exclude distribution of mothers education
black_em<-black_mobility[1:5, 3:7]

white_em<- white_mobility[1:5, 3:7]

#fertility for the birth matrix
black_GRR <- c(2.18, 2.18, 1.67, 1.43, 1.01)/ 6
white_GRR <- c(1.76, 1.81, 1.62, 1.54, 1.27)/ 6






```


#create matrices 

#Black survival matrix
```{r}

#notation- create a matrix that applies the rate of survival of each age group by educational attainment. This creates a matrix of survival of mothers 

b_edu<- unique(black_fem$group)

b_age<- as.numeric(unique(black_fem$age))

black_matrix_S <- list()
for(i in 2:length(b_age)) {
  new_matrix1 <- matrix(0, nrow=5, ncol=5) #inititalize a matrix
  survival <- c() #calculate surival with for loop
  for(j in 1:5) {
    survival <- c(survival, black_fem$nLx[which(black_fem$age==b_age[i] &  black_fem$group==b_edu[j])]/black_fem$nLx[which(black_fem$age==b_age[i]-5 & black_fem$group==b_edu[j])])
  }
  diag(new_matrix1) <- survival #apply surival rates to diagonal of matrix
  black_matrix_S <- c(black_matrix_S, list(new_matrix1)) #apply values to total matrix, creates a matrix for each education group at each age
  
}

print(black_matrix_S)

```

#white surivival matrx
```{r}

w_edu<- unique(white_fem$group) #collect unique education groups

w_age<- as.numeric(unique(white_fem$age)) #collect unique age groups

white_matrix_S <- list()
for(i in 2:length(w_age)) {
  new_matrix2 <- matrix(0, nrow=5, ncol=5) #initialize a matrix
  survival <- c() #calculate survival with for loop
  for(j in 1:5) {
    survival <- c(survival, white_fem$nLx[which(white_fem$age==w_age[i] &  white_fem$group==w_edu[j])]/white_fem$nLx[which(white_fem$age==w_age[i]-5 & white_fem$group==w_edu[j])])
  }
  diag(new_matrix2) <- survival #apply surival rates to diagonal of matrix
  white_matrix_S <- c(white_matrix_S, list(new_matrix2)) #apply values to total matrix, creates a matrix for each education group at each age
  
}

print(white_matrix_S)

```

#black birth matrix 

```{r}



black_matrix_B <- list()

for (i in 4:(length(b_age) - 1)) {
  new_matrix3 <- matrix(0, nrow = 5, ncol = 5)  # Initialize matrix
  survival <- c()  # Initialize survival of mothers and calculate

  for (j in 1:5) {
    survival <- c(
      survival,
      1 - 0.5 * (1 - black_fem$nLx[which(black_fem$age == b_age[i] & black_fem$group == b_edu[j])] /
                      black_fem$nLx[which(black_fem$age == b_age[i] - 5 & black_fem$group == b_edu[j])])
    )
  }

  new_matrix3 <- (black_em * black_GRR) * survival / 100  # Calculate mobility by GRR conditional on survival
  black_matrix_B <- c(black_matrix_B, list(new_matrix3))  # Apply values to matrix
} #and create a list 

print(black_matrix_B)
```

#create white birth matrix 

```{r}


white_matrix_B <- list()
for (i in 4:(length(w_age) - 1)) {
  new_matrix4 <- matrix(0, nrow = 5, ncol = 5)  # Initialize matrix
  survival <- c()  # Initialize survival of mothers and calculate
#iterate through each age group at reproductive age (10-45)
  for (j in 1:5) {
    survival <- c(
      survival,
      1 - 0.5 * (1 - white_fem$nLx[which(white_fem$age == w_age[i] & white_fem$group == w_edu[j])] /
                      white_fem$nLx[which(white_fem$age == w_age[i] - 5 & white_fem$group == w_edu[j])])
    )
  }

  new_matrix4 <- (white_em * white_GRR) * survival / 100  # Calculate mobility by GRR conditional on survival
  white_matrix_B <- c(white_matrix_B, list(new_matrix4))  # Apply values to matrix
} #and create a list 

print(white_matrix_B)
```

#create leslie matrix for black population

```{r}

#create dataframe of b elements
bb_dataframe <- as.data.frame(do.call(cbind, black_matrix_B))
bs_dataframe <- as.data.frame(do.call(cbind, black_matrix_S))


zeromatrix = matrix(0,5,5)

zeromatrix_df<-list()
zeromatrix_df<-c(zeromatrix_df, list(zeromatrix))
zeromatrix_df = as.data.frame(do.call(cbind, zeromatrix_df))

bb_row1<-c(zeromatrix_df,zeromatrix_df,zeromatrix_df, bb_dataframe, zeromatrix_df)
bb_row1_df<-as.data.frame(do.call(cbind, bb_row1))

bb_matrix<-as.matrix(bb_row1_df)


#create a dataframe for S elements
black_lms <- matrix(0, 45, 50)

for (i in 1:length(black_matrix_S)) {
  j <- 1:5 + 5*(i-1)
  black_lms[j, j] <- black_matrix_S[[i]]
}
black_lms


# x[1:5, 1:5] <- black_matrix_S[[1]]
# x[6:10, 6:10] <- black_matrix_S[[2]]

#combine birth matrix and survival matrix: 

black_lm <- rbind(bb_matrix, black_lms)

```


#create a leslie matrix for white population

```{r}

#create dataframe of b elements
wb_dataframe <- as.data.frame(do.call(cbind, white_matrix_B))

zeromatrix = matrix(0,5,5)

zeromatrix_df<-list()
zeromatrix_df<-c(zeromatrix_df, list(zeromatrix))
zeromatrix_df = as.data.frame(do.call(cbind, zeromatrix_df))

wb_row1<-c(zeromatrix_df,zeromatrix_df,zeromatrix_df, wb_dataframe, zeromatrix_df)
wb_row1_df<-as.data.frame(do.call(cbind, wb_row1))

wb_matrix<-as.matrix(wb_row1_df)


#create a dataframe for S elements
white_lms <- matrix(0, 50, 50)

for (i in 1:length(white_matrix_S)) {
  j <- 1:5 + 5*(i-1)
  white_lms[j, j] <- white_matrix_S[[i]]
}
white_lms


# x[1:5, 1:5] <- black_matrix_S[[1]]
# x[6:10, 6:10] <- black_matrix_S[[2]]

#combine birth matrix and survival matrix: 

white_lm <- rbind(wb_matrix, white_lms)


```

#create a starting population matrix based on provided education distribution of population in each age group. With the assumption that educational attainment is assigned at birth, the distribution of the population by education category provided in table 1 is used to distribute the population into education categories within each age group. 

#black population matrix

```{r}


# Given vector of values
black_ed <- c(37.6, 29.0, 23.3, 6.3, 3.8)

# Divide the vector by 100 and multiply by 1000
black_ed2 <- 1000 * (black_ed / 100)

# Initialize a 5 by 5 matrix
num_cols <- length(black_ed2)
num_repeats <- 10

# Repeat the process 10 times and bind the matrices
black_pop <- do.call(rbind, replicate(num_repeats, {
  black_pop <- matrix(0, nrow = num_cols, ncol = num_cols)
  diag(black_pop) <- black_ed2
  black_pop
}, simplify = FALSE))

# Print the result
print(black_pop)



```


#run the simulation for the black population 
```{r}

# Initialize a data frame to store the results
num_iterations <- 10000  # Adjust as needed
blackresult_df <- data.frame(matrix(0, nrow = dim(black_pop)[1], ncol = num_iterations))

# Set convergence threshold
convergence_threshold <- 1e-15

#transpose the matrix for multiplication
black_lm2 <- t(black_lm)

#normalize the data
oldblack_lm2 <- rowSums(black_pop)/sum(black_pop)

#to save distribution for each iteration create a list
biglistt <- list(oldblack_lm2)

#run the simulation, normalize the data within the interations
for(i in 1:num_iterations) {
  newblack_lm <- oldblack_lm2*black_lm2
  newblack_lm2 <- colSums(newblack_lm)
  newblack_lm3 <- newblack_lm2/sum(newblack_lm2)
 
#assign the new distribution to the list
  biglistt <- c(biglistt, list(newblack_lm3))
  if (all(abs(newblack_lm3 - oldblack_lm2) < convergence_threshold)) {
      break  # Exit the loop if convergence is met
  }
  oldblack_lm2 <- newblack_lm3
  print(i)
}

#distribution by education 
black_edu_dist <- matrix(NA, ncol=0, nrow=5)
for(i in 1:length(biglistt)) {
  time <- c(sum(biglistt[[i]][5*c(1:10)-4]), sum(biglistt[[i]][5*c(1:10)-3]), sum(biglistt[[i]][5*c(1:10)-2]), sum(biglistt[[i]][5*c(1:10)-1]), sum(biglistt[[i]][5*c(1:10)]))
 
  black_edu_dist <- cbind(black_edu_dist, time)
}





```

#create white population matrix 
```{r}


# Given vector of values
white_ed <- c(17.2, 21.9, 42.1, 11.9, 6.9)

# Divide the vector by 100 and multiply by 1000
white_ed2 <- 1000 * (white_ed / 100)

# Initialize a 5 by 5 matrix
num_cols <- length(white_ed2)
num_repeats <- 10

# Repeat the process 10 times and bind the matrices
white_pop <- do.call(rbind, replicate(num_repeats, {
  white_pop <- matrix(0, nrow = num_cols, ncol = num_cols)
  diag(white_pop) <- white_ed2
  white_pop
}, simplify = FALSE))

# Print the result
print(white_pop)

```
#run the simulation for the white population 
```{r}


# Initialize a data frame to store the results
num_iterations <- 10000  # Adjust as needed
whiteresult_df <- data.frame(matrix(0, nrow = dim(white_pop)[1], ncol = num_iterations))

# Set convergence threshold
convergence_threshold <- 1e-15

#transpose the matrix for multiplication
white_lm2 <- t(white_lm)

#normalize the data
oldwhite_lm2 <- rowSums(white_pop)/sum(white_pop)

#to save distribution for each iteration create a list
biglistt <- list(oldwhite_lm2)

#run the simulation, normalize the data within the interations
for(i in 1:num_iterations) {
  newwhite_lm <- oldwhite_lm2*white_lm2
  newwhite_lm2 <- colSums(newwhite_lm)
  newwhite_lm3 <- newwhite_lm2/sum(newwhite_lm2)
 
#assign the new distribution to the list
  biglistt <- c(biglistt, list(newwhite_lm3))
  if (all(abs(newwhite_lm3 - oldwhite_lm2) < convergence_threshold)) {
      break  # Exit the loop if convergence is met
  }
  oldwhite_lm2 <- newwhite_lm3
  print(i)
}

#distribution by education 
white_edu_dist <- matrix(NA, ncol=0, nrow=5)
for(i in 1:length(biglistt)) {
  time <- c(sum(biglistt[[i]][5*c(1:10)-4]), sum(biglistt[[i]][5*c(1:10)-3]), sum(biglistt[[i]][5*c(1:10)-2]), sum(biglistt[[i]][5*c(1:10)-1]), sum(biglistt[[i]][5*c(1:10)]))
 
  white_edu_dist <- cbind(white_edu_dist, time)
}


```


#create visualization for black education distribution 
```{r}

```

