---
title: "Problem Set 6"
author: "Sara Peters"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(kableExtra)
library(tinytex)
library(knitr)

#Set working directory
setwd("C:/Users/saraa/OneDrive - UW-Madison/SOC 756- Demography Techniques II/Problem Sets/DemTech2/Problem Set 6")

#load libraries
# install.packages("dplyr")
# install.packages("tidyverse")
# install.packages("ggplot2")
#install.packages("readxl")


library(dplyr)
library(tidyverse)
library(ggplot2)
#library (readxl)

# Set the 'scipen' option to a large value to prevent scientific notation
options(scipen = 300)

# Set the 'digits' option to control the number of decimal places
options(digits = 6)  # Change the number to the desired decimal places

```


###Soc 756

#Multiregional Projection

#This exercise requires that you adapt the model of Mare's 1997 "Differential Fertility…" article to the specific constraints and possibilities of the data supplied below. 

#You have the data given below in Tables 1-4 for black and white women for a given year. (Percentage totals in these tables may not sum to 100 because of rounding.)


#For this exercise, assume that (1) the population is closed to immigration and emigration, (2) the population has only one sex (females) and two race categories, (3) all fertility is confined between exact ages 15 and 45 and age-specific fertility rates are constant within this interval, and (4) fertility, mortality, and mobility rates are fixed over time. 

#Several of the questions below ask you to describe the equilibrium education distributions. This assumes that there is an equilibrium distribution of persons in the P(t+1) vector across education groups associated with the Leslie matrix and the starting P(0) vector. That is, repeated application of the Leslie matrix to the P(0) vector will eventually produce a P(n) vector in which the distribution of persons across education groups is not changing with each subsequent multiplication.



#This problem set will be submitted in two parts:


##Generate an M matrix with data for Black women and an M matrix with data for White women. You may have to make some decisions in constructing these. Make note of these for the answer to question number 1, to be submitted the following week. Upload your R script that constructs the two matrices to your GitHub page. The only “submission” this week to share the link to your uploaded file to me and two people in the class.




##Create Dataframes to supply matrices
```{r}

Lx_fem<- read.csv("nLx_values.csv")

#create separate black and white survival dataframes
Lx_fem$group <- substr(Lx_fem$cat_race_ed_age,1,3)

Lx_fem$race <- substr(Lx_fem$cat_race_ed_age, 2,2)

Lx_fem$age <- gsub(".*_", "", Lx_fem$cat_race_ed_age)

black_fem <- Lx_fem[which(Lx_fem$race=="b"),]

white_fem <- Lx_fem[which(Lx_fem$race=="w"),]

#create master mobility dataframes from Table 3 in mare
black_mobility <-data.frame(
  edu = c(8, 11, 12, 15, 16),
  black_edu = c(37.6, 29.0, 23.3, 6.3, 3.8), 
  black_em_8 = c(28.9, 7.9, 2.5, 3.3, 0.0), 
  black_em_11 = c(26.8, 35.0, 19.0, 3.8, 3.2), 
  black_em_12 = c(24.3, 27.8, 38.6, 24.3, 16.3), 
  black_em_15 = c(12.6, 16.4, 21.2, 49.6, 37.1), 
  black_em_16 = c(7.3, 12.9, 18.8, 18.9, 43.4) 
 
)

white_mobility <-data.frame(
  edu = c(8, 11, 12, 15, 16),
  white_edu = c(17.2, 21.9, 42.1, 11.9, 6.9),
  white_em_8 = c(13.2, 3.4, 1.4, 1.0, 0.1), 
  white_em_11 = c(17.9, 15.0, 6.1, 2.7, 3.3), 
  white_em_12 = c(48.5, 42.7, 45.7, 24.3, 14.3), 
  white_em_15 = c(13.0, 20.8, 24.0, 33.8, 25.9), 
  white_em_16 = c(7.5, 18.0, 21.7, 38.1, 56.4) 
)

#subset to exclude distribution of mothers education
black_em<-black_mobility[1:5, 3:7]

white_em<- white_mobility[1:5, 3:7]

#fertility for the birth matrix
black_GRR <- c(2.18, 2.18, 1.67, 1.43, 1.01)/ 6
white_GRR <- c(1.76, 1.81, 1.62, 1.54, 1.27)/ 6






```


#create matrices 

#Black survival matrix
```{r}

#notation- create a matrix that applies the rate of survival of each age group by educational attainment. This creates a matrix of survival of mothers 

b_edu<- unique(black_fem$group)

b_age<- as.numeric(unique(black_fem$age))

black_matrix_S <- list()
for(i in 2:length(b_age)) {
  new_matrix1 <- matrix(0, nrow=5, ncol=5) #inititalize a matrix
  survival <- c() #calculate surival with for loop
  for(j in 1:5) {
    survival <- c(survival, black_fem$nLx[which(black_fem$age==b_age[i] &  black_fem$group==b_edu[j])]/black_fem$nLx[which(black_fem$age==b_age[i]-5 & black_fem$group==b_edu[j])])
  }
  diag(new_matrix1) <- survival #apply surival rates to diagonal of matrix
  black_matrix_S <- c(black_matrix_S, list(new_matrix1)) #apply values to total matrix, creates a matrix for each education group at each age
  
}

print(black_matrix_S)

```

#white surivival matrx
```{r}

w_edu<- unique(white_fem$group) #collect unique education groups

w_age<- as.numeric(unique(white_fem$age)) #collect unique age groups

white_matrix_S <- list()
for(i in 2:length(w_age)) {
  new_matrix2 <- matrix(0, nrow=5, ncol=5) #initialize a matrix
  survival <- c() #calculate survival with for loop
  for(j in 1:5) {
    survival <- c(survival, white_fem$nLx[which(white_fem$age==w_age[i] &  white_fem$group==w_edu[j])]/white_fem$nLx[which(white_fem$age==w_age[i]-5 & white_fem$group==w_edu[j])])
  }
  diag(new_matrix2) <- survival #apply surival rates to diagonal of matrix
  white_matrix_S <- c(white_matrix_S, list(new_matrix2)) #apply values to total matrix, creates a matrix for each education group at each age
  
}

print(white_matrix_S)

```

#black birth matrix 

```{r}



black_matrix_B <- list()

for (i in 3:(length(b_age) - 1)) {
  new_matrix3 <- matrix(0, nrow = 5, ncol = 5)  # Initialize matrix
  survival <- c()  # Initialize survival of mothers and calculate

  for (j in 1:5) {
    survival <- c(
      survival,
      1 - 0.5 * (1 - black_fem$nLx[which(black_fem$age == b_age[i] & black_fem$group == b_edu[j])] /
                      black_fem$nLx[which(black_fem$age == b_age[i] - 5 & black_fem$group == b_edu[j])])
    )
  }

  new_matrix3 <- (black_em * black_GRR) * survival / 100  # Calculate mobility by GRR conditional on survival
  black_matrix_B <- c(black_matrix_B, list(new_matrix3))  # Apply values to matrix
} #and create a list 

print(black_matrix_B)
```

#create white birth matrix 

```{r}


white_matrix_B <- list()
for (i in 3:(length(w_age) - 1)) {
  new_matrix4 <- matrix(0, nrow = 5, ncol = 5)  # Initialize matrix
  survival <- c()  # Initialize survival of mothers and calculate
#iterate through each age group at reproductive age (10-45)
  for (j in 1:5) {
    survival <- c(
      survival,
      1 - 0.5 * (1 - white_fem$nLx[which(white_fem$age == w_age[i] & white_fem$group == w_edu[j])] /
                      white_fem$nLx[which(white_fem$age == w_age[i] - 5 & white_fem$group == w_edu[j])])
    )
  }

  new_matrix4 <- (white_em * white_GRR) * survival / 100  # Calculate mobility by GRR conditional on survival
  white_matrix_B <- c(white_matrix_B, list(new_matrix4))  # Apply values to matrix
} #and create a list 

print(white_matrix_B)
```

#create leslie matrix for black population

```{r}

#create dataframe of b elements
bb_dataframe <- as.data.frame(do.call(cbind, black_matrix_B))
bs_dataframe <- as.data.frame(do.call(cbind, black_matrix_S))


zeromatrix = matrix(0,5,5)

zeromatrix_df<-list()
zeromatrix_df<-c(zeromatrix_df, list(zeromatrix))
zeromatrix_df = as.data.frame(do.call(cbind, zeromatrix_df))

bb_row1<-c(zeromatrix_df,zeromatrix_df,bb_dataframe, zeromatrix_df)
bb_row1_df<-as.data.frame(do.call(cbind, bb_row1))

bb_matrix<-as.matrix(bb_row1_df)


#create a dataframe for S elements
black_lms <- matrix(0, 50, 50)

for (i in 1:length(black_matrix_S)) {
  j <- 1:5 + 5*(i-1)
  black_lms[j, j] <- black_matrix_S[[i]]
}
black_lms


# x[1:5, 1:5] <- black_matrix_S[[1]]
# x[6:10, 6:10] <- black_matrix_S[[2]]

#combine birth matrix and survival matrix: 

black_lm <- rbind(bb_matrix, black_lms)

```


#create a leslie matrix for white population

```{r}

#create dataframe of b elements
wb_dataframe <- as.data.frame(do.call(cbind, white_matrix_B))

zeromatrix = matrix(0,5,5)

zeromatrix_df<-list()
zeromatrix_df<-c(zeromatrix_df, list(zeromatrix))
zeromatrix_df = as.data.frame(do.call(cbind, zeromatrix_df))

wb_row1<-c(zeromatrix_df,zeromatrix_df,wb_dataframe, zeromatrix_df)
wb_row1_df<-as.data.frame(do.call(cbind, wb_row1))

wb_matrix<-as.matrix(wb_row1_df)


#create a dataframe for S elements
white_lms <- matrix(0, 50, 50)

for (i in 1:length(white_matrix_S)) {
  j <- 1:5 + 5*(i-1)
  white_lms[j, j] <- white_matrix_S[[i]]
}
white_lms


# x[1:5, 1:5] <- black_matrix_S[[1]]
# x[6:10, 6:10] <- black_matrix_S[[2]]

#combine birth matrix and survival matrix: 

white_lm <- rbind(wb_matrix, white_lms)


```

